Best Practices in Latent Class Analysis using Free Open Source Software

### 2023-03-24 manuscript.Rmd
change argument for BIC

---
title             : "Best Practices in Latent Class Analysis using the Free Open Source Software"
shorttitle        : "BEST PRACTICES LATENT CLASS ANALYSIS"
author: 
  - name          : "Caspar J. van Lissa"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address: "Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands"
    email: "c.j.vanlissa@tilburguniversity.edu"
    role:         # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
      - Conceptualization
      - Software
      - Supervision
      - Writing – original draft
      - Writing – review & editing
  - name          : "Mauricio Garnier-Villarreal"
    affiliation   : "2"
    corresponding : no   
    email         : "m.garniervillarreal@vu.nl"
    role:
      - Writing – review & editing
  - name          : "Daniel Anadria"
    affiliation   : "3"
    corresponding : no
    email         : "d.anadria@uu.nl"
    role:
      - Writing – original draft
      - Writing – review & editing
affiliation:
  - id            : "1"
    institution   : "Tilburg University, Methodology & Statistics"
  - id            : "2"
    institution   : "Vrije Universiteit Amsterdam, Sociology"
  - id            : "3"
    institution   : "Utrecht University, Methodology & Statistics"
author_note: |
  Methodology & Statistics
  Padualaan 14
  3582CH Utrecht, The Netherlands
abstract: |
  Latent class analysis (LCA) refers to a family of techniques for identifying groups
  in data based on a parametric model.
  Examples include mixture models, LCA with ordinal
  indicators, and latent class growth analysis.
  Despite the popularity of this technique, there is limited
  guidance with respect to best practices in estimating and reporting mixture
  models. Moreover, although user-friendly interfaces for advanced mixture
  modeling have long been available in commercial software packages, open
  source alternatives have remained somewhat inaccessible. This tutorial
  describes best practices for the estimation and reporting of LCA,
  using free and open source software in R. To this end, 
  we introduce new functionality for estimating and reporting mixture models
  using the `tidySEM` R-package.

@@

Finally, LCA can be used to deal with data which violate certain assumptions.
For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) [@baughman_mixture_2007; @magnus_multidimensional_2022].
These models estimate one class to explain the "true zeros",
and another class with a different distribution to explain all other values, which can include zero.
Note that this LCA has different models for both classes,
and thus falls beyond the scope of this paper.
<!-- CJ: This has been said elsewhere: -->
<!-- In fact, LCA assumes the population -->
<!-- distribution is a non-normal mixture of $k$ normal distributions, and it -->
<!-- can recover the value of $k$, i.e. generate a $k$-class solution. -->
<!-- CJ: This example is not perfect, we need another example still.. -->
For instance, a depression questionnaire PH-9 have found 2 distinct groups, ones with no symptoms, and one with symptoms. A mixture model can model the "presence" and "frequency" of symptoms as separate classes, instead of assuming that the sample follows a single continuous underlying distribution [@magnus_multidimensional_2022]. 
<!-- the aforementioned study on marine plastic noted that many prior studies had treated the distribution of particles as (multivariate) normal, despite clear deviations from normality [@alkema_maximizing_2022].
A mixture of two normal distributions with distinct variances described the observed distribution better. -->
<!-- LCA and factor analysis share the approximation of an unobserved latent variable, the difference being if that variable is continuous or categorical. We encourage researchers to be agnostic about the **true** nature of the latent variable. And focus on the approach that fits the research question. -->
<!-- CJ: Why would we encourage researchers to be agnostic about the true...? -->

@@

Since SA is a global optimizer, it should be independent of starting values.
With this in mind, it is efficient to start estimation from a "reasonable solution", instead of a random starting point.
For example, if we assume that the different classes are likely to have different mean values on the indicators, 
then a nonparametric clustering algorithm - like K-means or hierarchical clustering - can be used to determine these cluster centroids [see also @biernacki_choosing_2003].
By default, `tidySEM` derives starting values using K-means clustering [@likas_global_2003; @Forgy1965ClusterAO; @Hartigan_kmeans],
computing the mixture model starting values by treating the K-means solution as multi-group model.
Next, SA is used to find the global optimum solution.
Finally, SA is followed up with a short run of the ML algorithm,
as ML inherently produces an asymptotic
covariance matrix for the parameters that can be used to compute
standard errors.
Note that these defaults can be manually overridden.

@@

Due to the lack of absolute fit indices for LCA, the assessment of relative fit
relies on information criteria.
Information criteria (ICs) are computed from the $-2LL$,
which is the most basic fit measure, testing the overall hypothesis if the model properly represents the data.
In addition to capturing model fit, ICs add a penalty for model complexity thus
preferring more model fit and parsimony.
The lower the value of an information criterion, the better the overall fit of the
model.
The *Akaike Information Criterion* (*AIC*) is the original information criterion.
It is computed as $-2LL+2p$, where $p$ is the number of parameters.
The *Bayesian Information Criterion* (*BIC*) multiplies the penalty for complexity with a function of the sample size, $p \ln(n)$,
such that complex models are penalized more in smaller samples.
Finally,
the so-called *sample size adjusted BIC* (*saBIC*) implements a different ad-hoc penalty for sample size,
$\ln(\frac{n + 2}{24})$. Although simulation studies have shown that the saBIC performs well for class enumeration [@tein_statistical_2013; @chen_efficacy_2017],
a limitation is that its penalty is not based on theory, but on the Minimum Description Length (MDL) as the shortes set of information necessary to reproduce the data [@Rissanen_1985; @sclove_application_1987]. BIC also performs well and is theoretically substantiated.
As such, BIC may be the most appropriate information criterion to use for model comparison [@nylund-gibson_ten_2018; @masyn_latent_2013]. 
In `tidySEM`, all three ICs are available as a part of the output of `table_fit()`.

@@

The function `table_fit()` reports an overview of fit indices, customized for LCA.
In reporting model fit, it is important, firstly, to report the effective sample size, which should be the same for all models in order to compare relative fit measures.
Secondly, it should be clearly reported,
for example, in the strategy of analyses,
which parameters are estimated across and within classes.
In case of doubt, a full overview of the parameters is given by the `coef()` and `table_results()` functions.
The number of parameters for each model should be included in the fit table.
Thirdly, it is important that readers can assess the relative fit of different solutions.
The most fundamental fit measure is the $-2LL$,
as this is the quantity optimized during estimation.
From these statistics, information criteria can be computed.
By default, `table_fit()` includes the AIC, BIC, and saBIC.
We suggest to report at least the BIC, as it is ubiquitous, grounded in theory, and performs well in class enumeration.

>>>>

---
title             : "Best Practices in Latent Class Analysis using the Free Open Source Software"
shorttitle        : "BEST PRACTICES LATENT CLASS ANALYSIS"
author: 
  - name          : "Caspar J. van Lissa"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address: "Professor Cobbenhagenlaan 125, 5037 DB Tilburg, The Netherlands"
    email: "c.j.vanlissa@tilburguniversity.edu"
    role:         # Contributorship roles (e.g., CRediT, https://credit.niso.org/)
      - Conceptualization
      - Software
      - Supervision
      - Writing – original draft
      - Writing – review & editing
  - name          : "Mauricio Garnier-Villarreal"
    affiliation   : "2"
    corresponding : no   
    email         : "m.garniervillarreal@vu.nl"
    role:
      - Writing – review & editing
  - name          : "Daniel Anadria"
    affiliation   : "3"
    corresponding : no
    email         : "d.anadria@uu.nl"
    role:
      - Writing – original draft
      - Writing – review & editing
affiliation:
  - id            : "1"
    institution   : "Tilburg University, Methodology & Statistics"
  - id            : "2"
    institution   : "Vrije Universiteit Amsterdam, Sociology"
  - id            : "3"
    institution   : "Utrecht University, Methodology & Statistics"
author_note: |
  Methodology & Statistics
  Padualaan 14
  3582CH Utrecht, The Netherlands
abstract: |
  Latent class analysis (LCA) refers to a family of techniques for identifying groups
  in data based on a parametric model.
  Examples include mixture models, LCA with ordinal
  indicators, and latent class growth analysis.
  Despite the popularity of this technique, there is limited
  guidance with respect to best practices in estimating and reporting mixture
  models. Moreover, although user-friendly interfaces for advanced mixture
  modeling have long been available in commercial software packages, open
  source alternatives have remained somewhat inaccessible. This tutorial
  describes best practices for the estimation and reporting of LCA,
  using the free and open source software in R. To this end, 
  we introduce new functionality for estimating and reporting mixture models
  using the `tidySEM` R-package.

@@

Finally, LCA can be used to deal with data which violate certain assumptions.
For example, zero-inflated and hurdle models are special cases of LCA for dealing data with a very high proportion of zeros (extreme right skew) [@baughman_mixture_2007].
These models estimate one class to explain the "true zeros",
and another class with a different distribution to explain all other values, which can include zero.
Note that this LCA has different models for both classes,
and thus falls beyond the scope of this paper.
<!-- CJ: This has been said elsewhere: -->
<!-- In fact, LCA assumes the population -->
<!-- distribution is a non-normal mixture of $k$ normal distributions, and it -->
<!-- can recover the value of $k$, i.e. generate a $k$-class solution. -->
<!-- CJ: This example is not perfect, we need another example still.. -->
One example of mixture models for violations of normality 
is in modeling responses to the PH-9, a questionnaire of depressive symptoms.
Researchers observed two distinct groups: One with no symptoms on most indicators (zero inflation),
and one with non-zero symptoms [@magnus_multidimensional_2022].
As this response pattern violates the assumption of multivariate normality,
the authors used a mixture model to capture the "presence" and "frequency" of symptoms as separate classes.
<!-- the aforementioned study on marine plastic noted that many prior studies had treated the distribution of particles as (multivariate) normal, despite clear deviations from normality [@alkema_maximizing_2022].
A mixture of two normal distributions with distinct variances described the observed distribution better. -->
<!-- LCA and factor analysis share the approximation of an unobserved latent variable, the difference being if that variable is continuous or categorical. We encourage researchers to be agnostic about the **true** nature of the latent variable. And focus on the approach that fits the research question. -->
<!-- CJ: Why would we encourage researchers to be agnostic about the true...? -->

@@

Since SA is a global optimizer, it should be independent of starting values.
With this in mind, it is efficient to start estimation from a "reasonable solution", instead of a random starting point.
For example, if we assume that the different classes are likely to have different mean values on the indicators, 
then a nonparametric clustering algorithm - like K-means or hierarchical clustering - can be used to determine these cluster centroids [see also @biernacki_choosing_2003].
By default, `tidySEM` derives starting values using K-means clustering [@Hartigan_kmeans],
computing the mixture model starting values by treating the K-means solution as multi-group model.
Next, SA is used to find the global optimum solution.
Finally, SA is followed up with a short run of the ML algorithm,
as ML inherently produces an asymptotic
covariance matrix for the parameters that can be used to compute
standard errors.
Note that these defaults can be manually overridden.

@@

Due to the lack of absolute fit indices for LCA, the assessment of relative fit
relies on information criteria.
Information criteria (ICs) are computed from the $-2LL$,
which is the most basic fit measure, testing the overall hypothesis if the model properly represents the data.
In addition to capturing model fit, ICs add a penalty for model complexity thus
preferring more model fit and parsimony.
The lower the value of an information criterion, the better the overall fit of the
model.
The *Akaike Information Criterion* (*AIC*) is the original information criterion.
It is computed as $-2LL+2p$, where $p$ is the number of parameters.
The *Bayesian Information Criterion* (*BIC*) multiplies the penalty for complexity with a function of the sample size, $p \ln(n)$,
such that complex models are penalized more in smaller samples.
Finally,
the so-called *sample size adjusted BIC* (*saBIC*), $\ln(\frac{n + 2}{24})$,
implements a penalty for sample size that is based on the Minimum Description Length, a concept from data compression that refers to the smallest number of pieces of information necessary to reproduce the data [@Rissanen_1985; @sclove_application_1987].
Simulation studies have shown that the saBIC performs well for enumeration of homogenous clusters in small samples,
but tends to select too many classes when the number of the number of classes and the sample size are large [@tein_statistical_2013; @chen_efficacy_2017].
The BIC has been found to perform consistently well across simulation studies,
and is among the most widely known information criteria.
As such, BIC may be a sensible default criterion for model comparison [@nylund-gibson_ten_2018; @masyn_latent_2013]. 
In `tidySEM`, all three ICs are available as a part of the output of `table_fit()`.

@@

The function `table_fit()` reports an overview of fit indices, customized for LCA.
In reporting model fit, it is important, firstly, to report the effective sample size, which should be the same for all models in order to compare relative fit measures.
Secondly, it should be clearly reported,
for example, in the strategy of analyses,
which parameters are estimated across and within classes.
In case of doubt, a full overview of the parameters is given by the `coef()` and `table_results()` functions.
The number of parameters for each model should be included in the fit table.
Thirdly, it is important that readers can assess the relative fit of different solutions.
The most fundamental fit measure is the $-2LL$,
as this is the quantity optimized during estimation.
From these statistics, information criteria can be computed.
By default, `table_fit()` includes the AIC, BIC, and saBIC.
We suggest to report at least the BIC, as it is ubiquitous, well-understood,
and performs well in class enumeration.
